### 2019-03-15 paper/sex_selection_spacing-ver2.tex
Write-up of distribution results

Average spacing is a convenient way to understand the overall changes but may hide 
important differences in the distribution of spacing.
Specifically, a concern is that the increases in average spacing that I show above may 
mask an increased variance in birth spacing with families with fewer sons both more likely 
to have a very pregnancy and to continue through pregnancies and sex-selective abortions 
until they conceive a son.
This section, therefore, provides more detail on how spacing is distributed within a spell 
and across sex compositions using a graphical approach.

I show survival curves conditional on predicted parity progression rather than standard 
survival curves. 
The advantage of this approach is that it is possible to directly compare the distribution 
of spacing to next birth across groups, independently of differences in how likely the 
next birth is.
Because the conditional survival curves are independent of the likelihood of parity 
progression within the spell, they all begin at 100\% and end at 0\%.

@@

For women with no education, Figure \ref{fig:pps_low} shows that the median results
above are consistent with the distribution of spacing within a spell with spacing 
after only girls consistently shorter than after only boys.
There are slight variations in how strong this effect is across the three spells.
On the one hand, for the second spell the difference in spacing between
when the first child is a girl versus when the first child is a boy as
first-born remains almost the same across the four decades.
On the other hand, for the third spell the differences between two girls and 
the other sex compositions become more pronounced over time, and especially
the difference between two girls and two boys have become substantially larger.
The fourth spell results are based on relatively few observations but do
indicate that women with either only girls or two girls and one boy have
shorter spacing than women with only boys or one girl and two boys.
Although the evidence is not substantial, it does appear that there is a slight
preference for two boys in the sense that spacing becomes longer once the couple
has at least two boys than for the other sex compositions.

@@

The distribution of spacing for the second spell is almost indistinguishable 
between urban women with eight or more years of education and rural women 
with no education for the 1972--1984 period, as shown in Figure \ref{fig:pps_high}.
The small difference points to these two very different groups of women behaving similarly
in response to son preference in a situation where there was little access to
prenatal sex determination or little incentive to use it early even for those 
who might have had access.

Whereas the no education group show almost no change across the four decades,
the high education women show a rightward shift if the spacing pattern when
the first child is a girl.
The result is that the spacing patterns are close to identical after first-born 
girls and after first-born boys.
Given the elevated predicted percentage boys in Table \ref{tab:median_sex_ratio_high},
this change is consistent with the use of sex selection for women with a girl 
as their first child.

The hypothesized correlation between use of sex selection and reversal of 
the standard spacing pattern show up clearly in the 2005--2016 period for the 
third spell with spacing to the third child consistently longer with two girls 
than with either two girls or one boy and one girl.
The rightward movement for only girls from the, mostly, standard pattern in 
the 1972--1984 period to the new pattern in the 2005--2016 period is substantial.
There are also indications that the spacing distribution has moved rightward
for women with one boy and one girl relatively to women with two boys, which
provides some evidence for the use of sex selection for this group.
The caveat is that the 1972--1985 period also had a biased sex ratio, which
points to uncorrected recall error for this group.

There is an even more pronounced rightward shift for spacing after only
girls in the fourth spell for the first years of the spell. 
The caveat is that there are few births to base the results on,
especially for women with two or more sons.%
\footnote{
As mentioned, it is possible that some women had access to sex selection even
in the first period.
I would expect well educated urban women with more children to be the
most likely to know about and use this early access to prenatal sex determination.
}
Furthermore, the differences are substantial across most of the distribution.
Twelve months after the beginning of the spell there is a more than ten 
percentage points difference in the conditional survival curves for only girls
and only boys.
That is, of the women who are predicted to have a fourth child just over 30\%
of women with three boys have had their fourth child, while less than 20\% of
those with three girls have had theirs.
Even at three years after the beginning of the spell, there is a more than 
five percentage points difference.

As discussed above there is some evidence of a two-son preference with
the spacing pattern almost the same for women with three boys and with
two boys and one girl.
The conditional survival curve for women with one boy and two girls is
consistently above those for women with two or more son, but below those 
with three girls.
The pattern fit the predicted sex ratios shown above where the highest
predicted sex ratio is for women with only girls, followed by those
with one son and two girls, and those with two or more boys having
almost the same sex ratio.

@@

The West contains states with a high degree of son preference and use of sex selection.
The North consists of states that also have a strong son preference, but initially had
higher fertility and therefore lower likelihood of using sex selection.
The East states generally have only a moderate level of son preference.
Finally, the South contains states with traditionally low son preference and lower
fertility than the rest of India.

>>>>

Average spacing is a convenient way to understand the overall changes but may hide 
important differences in the distribution of spacing.
Specifically, a concern is that the increases in average spacing that I find above may 
mask an increased variance in birth spacing for families with fewer sons if they are both 
more likely to have a very early pregnancy and to continue through pregnancies and 
sex-selective abortions until they conceive a son.
This section, therefore, provides more detail on how spacing is distributed within a spell 
and across sex compositions using a graphical approach.

I show survival curves conditional on predicted parity progression rather than standard 
survival curves. 
The advantage of this approach is that it is possible to directly compare the distribution 
of spacing to next birth across groups, independently of differences in how likely the 
next birth is.
Because the conditional survival curves are independent of the likelihood of parity 
progression, they all begin at 100\% and end at 0\%.

@@

Figure \ref{fig:pps_low} shows the distribution of birth within a spell for women with no 
education.
For the second spell, what is most interesting is that there is almost no difference 
between those with a first-born boy and those with a first-born girl.
For the 1972--1984 period the two conditional survival curves only begin to diverge 
at around 18 months (corresponding to 27 months after the first birth took place), and
at that point close to 50 percent of the women have already had their second child.
The literature on birth spacing and child health suggests that an interpregnancy interval
of less than 18 months is associated with a higher risk of adverse health outcomes and
mortality \citep{Conde-Agudelo2012}.
With a nine-month pregnancy this means that almost 50 percent of the children born fall
in this category, and, somewhat surprisingly, that there do not appear to be any particular 
disadvantage for the child born after a girl than after a boy.%
\footnote{
Note that this does not imply that the first-born girl is not in a more adverse position
than a first-born boy when the second child arrives.
}
The difference in spacing between a first-born girl and a first-born boy first-born 
remains almost the same across the four decades.
Furthermore, the proportion with very short spacing remains almost the same.

For the third spell the son preference driven short spacing is evident throughout the
distribution and becomes more pronounced over the four decades.
Families with two girls are especially likely to have a birth early with 25 percent
having a child within 12 months of the beginning of the spell (21 months after the prior
second birth).
That number is only slightly smaller four decades later. 
The fourth spell shows changes in the distribution that appear to be more pronounced
than for the third spell, although these results are based on fewer observed births.
For families with three girls, 75 percent have had their fourth child within 36 months 
after the beginning of the spell, but this occurs more than a year later for families
with two or more boys.

@@

The distribution of spacing for the second spell is almost indistinguishable 
between urban women with eight or more years of education and rural women 
with no education for the 1972--1984 period, as shown in Figure \ref{fig:pps_high}.
The small differences point to these two very different groups of women behaving similarly
in response to son preference in a situation where there was little access to
prenatal sex determination or little incentive to use it early even for those 
who might have had access.
However, whereas the no education group shows almost no change across the four decades for 
the second spell, the high education women show an upward shift and a relatively even
distribution of births within the spell for both women with a first-born son and women
with a first-born girl.
The result is that the spacing patterns are close to identical after first-born 
girls and after first-born boys until after approximately 80 percent have had their 
second child, although as mentioned above, this is driven in large part by the significant
use of sex selection among women with a first-born girl.

The correlation between the increased use of sex selection and the reversal of the 
standard spacing pattern show up clearly in the 2005--2016 period for the third spell with 
spacing to the third child consistently longer with two girls than with either two boys 
or one boy and one girl throughout the entire spell.%
\footnote{
For the third spell, the 1972--1984 period show a close to standard pattern throughout, 
although, as mentioned above, the sex ratios for women without girls was substantially 
higher than the natural sex ratio either because of recall error or early use of 
sex selection.
}
There is an even more pronounced rightward and upward shift for spacing after only
girls in the fourth spell for the first years of the spell.%
\footnote{
The caveat is that there are few births to base the results on,
especially for women with two or more sons.
} 
Furthermore, the differences are substantial across most of the distribution.
Twelve months after the beginning of the spell there is a more than ten 
percentage points difference in the conditional survival curves for only girls
and only boys.
Both of the spells therefore show changes in the distribution of births that is in line
with extensive use of sex selection among families with no or one son.

% Even with sex selection, however, the observed pattern for the third and fourth spells
% is not preordained.
% We should expect 51.2 percent of the women to conceive a boy when they have their first 
% pregnancy in the spell, which means that they do not need to use sex selection to have
% a son.
% Hence, even with sex selection, we might expect to observe a very short interval for the 
% initial births if families with strong son preference still try to conceive sooner than 
% those with lower son preference.
% That we do not observe a short intervals for the early births calls into question the 
% assumption that son preference leads to very early births with fewer sons even in 
% situations when there is access to sex selection.
% If, however, parents know that children born after short interpregnancy periods suffer 
% from worse health outcomes it makes sense to wait longer before conceiving again in
% this situation, precisely because they know that with sex selection the next child they 
% carry to term will be a boy.
% 
% Hence, there are two potential reasons why son preference combined with access to
% sex selection can lead to longer birth intervals.
% First, the ``abortion'' effect, where each abortion extend birth spacing by between six
% and twelve months.
% Second, a ``delay'' effect, where parents conceive later because they know that short
% spacing is detrimental to the next child's health, and that the next child is more likely
% to be a boy because of access to sex selection.
% The ``delay'' effect may explain why the twenty-five percentile birth intervals for the
% second spell increase more for families with two girls than either of the other two
% sex compositions, even though there should be few if any sex-selective abortions among
% these births.%
% \footnote{
% See Appendix Table \ref{tab:p25_p50_p75_high}.
% }

@@

The ``West'' contains states with a high degree of son preference and use of sex selection.
The ``North'' consists of states that also have a strong son preference, but initially had
higher fertility and therefore lower likelihood of using sex selection.
The ``East'' states generally have only a moderate level of son preference.
Finally, the ``South'' contains states with traditionally low son preference and lower
fertility than the rest of India.
